---
title: "Early vertebrates dispRity pipeline"
author: "RPDearden"
date: "06/11/2018"
output: rmarkdown::github_document
---

This is me having a play with the functionalities of dispRity, using the matrix from [Keating *et al.* 2016](http://rspb.royalsocietypublishing.org/content/283/1826/20152917).  The tree is a majority rule consensus I obtained by running the dataset in PAUP*, and the dates are cobbled together from the literature.

###Preliminaries###

Read in required packages

```{r message=FALSE, warning=FALSE}
library(dispRity)
library(geomorph)
library(vegan)
library(cluster)
library(Claddis)
```

Set directory and read in nexus file - NOTE ReadMorphNexus reads in both ? and - as NA - instead I have read the matrix in as a csv file (by copying and pasting from Mesquite because I am basic)

```{r warning=FALSE}
setwd("~/Documents/Side_projects/Masters_writeup/EarlyVerts_R_Code/dispRity_Pipeline")
#Kdata <- ReadMorphNexus("Keating16.nex")
Ktree <- read.nexus("KeatMaj.tre")
Kdates <- read.csv("Keating16Dates.csv", header=TRUE,row.names=1)
Kmatrix <- read.csv("Keating16.csv", header=TRUE, row.names=1, stringsAsFactors = FALSE, na.strings="?") #Reads ? as NA?
```

Change values as detailed in that Deline paper (find)
```{r warning=FALSE}
#Replace - with -1 (will become 0 later)
Kmatrix[Kmatrix=="-"] <- -1
#Convert all remaining character columns into numeric
indx <- sapply(Kmatrix, is.character)
Kmatrix[indx] <- lapply(Kmatrix[indx], function(Kmatrix) as.numeric(as.character(Kmatrix))) 
 #Add 1 to all values
Kdata <- Kmatrix+1

```

Ordinate - this is the NMDS procedure I used in original project - possible to refine?
```{r warning=FALSE}
Kdistances=daisy(Kdata, metric=c("gower"))
kdataNMDS<-metaMDS(Kdistances, k=3, zerodist="add")
#n.b. Ordination is stored in:
Ordkdata<-kdataNMDS$points
```

Plot first two axes of NMDS - plot looks like the back end of a bus but not much point twiddling currently
```{r warning=FALSE}
#Normal with text
plot(kdataNMDS)
text(kdataNMDS, cex=0.8)
```

Test stresses
```{r warning=FALSE, results='hide'}
#Run metaMDS with different numbers of axes
NMDS1=metaMDS(Kdistances, k=1, zerodist="add")
NMDS2=metaMDS(Kdistances, k=2, zerodist="add")
NMDS3=metaMDS(Kdistances, k=3, zerodist="add")
NMDS4=metaMDS(Kdistances, k=4, zerodist="add")
NMDS5=metaMDS(Kdistances, k=5, zerodist="add")
NMDS6=metaMDS(Kdistances, k=6, zerodist="add")
NMDS7=metaMDS(Kdistances, k=7, zerodist="add")
NMDS8=metaMDS(Kdistances, k=8, zerodist="add")
NMDS9=metaMDS(Kdistances, k=9, zerodist="add")
NMDS10=metaMDS(Kdistances, k=10, zerodist="add")

#Make vector containing stresses
Stresses=c(NMDS1$stress, NMDS2$stress, NMDS3$stress, NMDS4$stress, NMDS5$stress, NMDS6$stress, NMDS7$stress, NMDS8$stress, NMDS9$stress, NMDS10$stress)
#Plot it
barplot(Stresses, ylim=c(0, 0.30), names.arg=c(1:10), xlab="No. of axes", ylab="Stress")
```


<!-- Alternatively originally did like below lazily using default Claddis.ordination settings, but have a think about appropriate methods -->
<!-- ```{r warning=FALSE} -->
<!-- OrdKdata <- Claddis.ordination(Kdata) -->
<!-- ``` -->

###Doing disparity through time (disparity metric used the sum of variances as default)###

tree requires a root time - for now guessed at start of Cambrian, choose a better one
```{r warning=FALSE}
Ktree$root.time <- 541
```

Order dates columns to match tip labels in tree (necessary for Claddis{DatePhylo})
```{r warning=FALSE}
Kdates <- Kdates[match(Ktree$tip.label, rownames(Kdates)),]
```

Date tree using Claddis tree dating - check dating method appropriate
```{r warning=FALSE}
DatedKtree<-DatePhylo(Ktree, Kdates, method="equal", rlen=1)
```

Make time bins (random 50my)
```{r warning=FALSE}
time_bins <- rev(seq(from = 250, to = 600, by = 50))
```

Split dataset into subsets
```{r warning=FALSE}
binned_Kdata <- chrono.subsets(data = Ordkdata, tree = Ktree, method = "discrete", time = time_bins, inc.nodes = FALSE,FADLAD = Kdates)
```

Bootstrapping , first just that and then with bonus rarefaction
```{r warning=FALSE}
boot_bin_Kdata <- boot.matrix(binned_Kdata)
rare_bin_Kdata <- boot.matrix(binned_Kdata, bootstraps = 100,rarefaction = 6)
```

Calculate disparity for both datasets
```{r warning=FALSE}
boot_disparity_Kdata <- dispRity(boot_bin_Kdata, metric = c(sum, variances))
rare_disparity_Kdata <- dispRity(rare_bin_Kdata, metric = c(sum, variances))
```

Plot it up
```{r warning=FALSE}
quartz(width = 15, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
plot(boot_disparity_Kdata, type = "continuous", main = "bootstrapped results")
plot(rare_disparity_Kdata, type = "continuous", main = "rarefied results")
```

<!-- Testing bins difference - won't work with this dataset as not enough data -->
<!-- ```{r warning=FALSE} -->
<!-- #test.dispRity(boot_disparity_Kdata, test = wilcox.test, comparison = "sequential",correction = "bonferroni") -->
<!-- ``` -->
