---
title: "Character rates pipeline"
author: "RPDearden"
date: "07/11/2018"
output: rmarkdown::github_document
---

This is me having a play with the character rate functionalities of Claddis, using the matrix from [Keating *et al.* 2016](http://rspb.royalsocietypublishing.org/content/283/1826/20152917).  The tree is a majority rule consensus I obtained by running the dataset in PAUP*, and the dates are cobbled together from the literature.

###Preliminaries###

Read in required packages

```{r message=FALSE, warning=FALSE}
library(Claddis)
library(paleotree)
library(strap)
library(ape)
library(plotrix)
```

Set directory and read in nexus file - NOTE ReadMorphNexus reads in both ? and - as NA - instead I have read the matrix in as a csv file (by copying and pasting from Mesquite because I am basic)

```{r warning=FALSE}
setwd("~/Documents/Side_projects/Masters_writeup/EarlyVerts_R_Code/CharRates_pipeline")
Kdata <- ReadMorphNexus("../Keating16.nex")
Ktree <- read.nexus("../KeatMaj.tre")
Kdates <- read.csv("../Keating16Dates.csv", header=TRUE,row.names=1)
tbins<-c(541,509,497,485.4,477.7,470,458.4,443.4,433.4,419.2,410.8,393.3,382.7,372.2,358.9,346.7,330.9,323.2,307,298.9,290.1,272.3,259.8,252.17)
```

Prepare tree - think about how break polytomies (either many random iterations or by time of branching in paleotree)).  Also think about best way to date tree - better from molecular info?
```{r warning=FALSE}
#Break polytomies randomly
Ktree <- multi2di(Ktree)
#Reorder dates file to get in line with tree file
Kdates <- Kdates[match(Ktree$tip.label, rownames(Kdates)),]
#Scale tree using fads and lads
Ktree <- DatePhylo(Ktree, Kdates,method="equal", rlen=1)
#plot with geoscale (to check worked)
geoscalePhylo(ladderize(Ktree), cex.age=0.6, cex.ts=0.8, cex.tip=0.8)
```

<!-- #Prune above to get in line with daterr MAY NOT NEED THIS BUT JUST IN CASE-->
<!-- nms.id <- complete.cases(daterr)				# labels the taxa with complete data -->
<!-- daterr <- as.matrix(daterr[nms.id,]) 		# saves the labelled taxa -->
<!-- nms.id <- complete.cases(ages.tree)				# labels the taxa with complete data -->
<!-- drop.names <- rownames(ages.tree)[!nms.id]			# switches labelling of TRUE and FALSE #switched up from below to fix -->
<!-- ages.tree <- as.matrix(ages.tree[nms.id,]) 		# saves the labelled taxa -->
<!-- tree.data <- drop.tip(tree.data, drop.names)	# returns a list of taxa with complete data -->

###Plotting rates on branches and nodes###
Discrete character rates with Claddis
```{r warning=FALSE}
#Calculate rates
#rate.data <- DiscreteCharacterRate(Ktree, Kdata, seq(Ktree$root.time, Ktree$root.time - max(diag(vcv(Ktree))), length.out=6), alpha=0.01)

#With my time bins - check bins appropriate 
rate.data <- DiscreteCharacterRate(Ktree, Kdata, tbins, alpha=0.01)
```
Plot with edges
```{r warning=FALSE}
#Set up edge colours
edge.color <- rep("black", nrow(Ktree$edge))
edge.color[which(rate.data$branch.results[, "ml.signif.hi"] == 1)] <- "red"
edge.color[which(rate.data$branch.results[, "ml.signif.lo"] == 1)] <- "blue"

#Plot with these colours
# pdf("VertEdgeRates.pdf")
geoscalePhylo(ladderize(Ktree), cex.age=0.6, cex.ts=0.8, cex.tip=0.8, edge.color=edge.color[match(ladderize(Ktree)$edge[, 2], Ktree$edge[,2])])
# dev.off()
```

Plot for nodes
```{r warning=FALSE}
#Set up node colours
node.color <- rep("white", nrow(rate.data$node.results))
node.color[which(rate.data$node.results[, "ml.signif.hi.ib"] == 1)] <- "red"
node.color[which(rate.data$node.results[, "ml.signif.lo.ib"] == 1)] <- "blue"
node.color[which(is.na(rate.data$node.results[, "ml.signif.lo.ib"]))] <- NA

# Plot tree with node rates
#pdf("VertNodeRates.pdf")
geoscalePhylo(Ktree, cex.age=0.6, cex.ts=0.8, cex.tip=0.8)
nodelabels(node=rate.data$node.results[, "node"][!is.na(node.color)], pch=21, col="black", bg=node.color[!is.na(node.color)])
#dev.off()
```

###Plotting rates through time###

The following is the procedure I used in my Masters, adapted from Lloyd (2012)
```{r warning=FALSE}
source("http://www.graemetlloyd.com/pubdata/functions_2.r")
# Load tree: n.b. this has to have branch lengths - I got ACCTRAN and DELTRAN optimised trees from paup by using set=ACCTRAN/DELTRAN and then then saving a tree with branch lengths (BrLens=yes)
tree <- read.nexus("../KeatMajACC.tre")
# Load ages:
ages<-daterr<-read.csv("../Keating16Dates.csv", header=TRUE,row.names=1)
# Load completeness (used in one of the dating methods):
comp<-read.csv("../K16Completeness.csv")          # number of missing characters - for now just counted ?
#Randomise polytomies
tree<-multi2di(tree, random=TRUE)

# Number of characters (used in one of the dating methods):
nchar<-120
# Set number of trials:
ntrials <- 1000 #down from 1000
# Get list of nonterminal and nonroot nodes (Ntips plus two gets first non-root node, second is all nodes):
nodes<-(Ntip(tree)+2):(Ntip(tree)+Nnode(tree))

# Calculate tree properties
#How many branches?
n <- length(tree$edge[,1])
#Creates a vector of edge lengths for the tree
changes <- tree$edge.length
# Create patristic distance tree -> function in Claddis? (patristic - distance between two tips)
ptree<-pat.dist.phylo(tree, comp, nchar)

# Calculate additional tree properties
comparablechars <- changes/ptree$edge.length   # non-missing characters
comparablechars[grep(NaN,comparablechars)] <- 0 # Catch zero changes/zero time; modified!
pctcomp <- comparablechars/nchar
#MakeTimeBins
#tbins<-c(541,358.9,346.7,330.9,323.2,307,298.9,290.1,272.3,259.8,252.17)
tbins <- c(541,307,250)
# X-axis for plots, finds middles of time bins:
#Changed first + to a -, think this is correct
x<-tbins[1:(length(tbins)-1)]-((tbins[1:(length(tbins)-1)]-tbins[2:length(tbins)])/2)
# Make matrix for storing counts:
acctran.sumlineages<-acctranbins<-matrix(ncol=ntrials,nrow=length(tbins)-1)

# Main loop:
for (i in 1:ntrials) {
   # Pick a random edge between the limits: (for each column pics a random age from a uniform distribution between the FAD and LAD)
    for (j in 1:length(ages[,1])) ages[j,1]<-runif(1,daterr[j,2],daterr[j,1])
    #dates phylogeny on basis of these dates
    ttreeb<-date.phylo(tree, ages[tree$tip.label,], rlen=2, method="ruta", ptree)
    nodeages<-date.nodes(ttreeb, max(ages)+2)
    los<-nodeages[ttreeb$edge[,2]]
    fos<-nodeages[ttreeb$edge[,1]]
    changes<-vector(mode="numeric")
    for (j in 1:length(tree$edge.length))       changes<-c(changes,runif(tree$edge.length[j],nodeages[ttreeb$edge[j,2]],nodeages[ttreeb$edge[j,1]]))
    for (j in 2:length(tbins)) {
        acctranbins[(j-1),i]<-length(intersect(grep(TRUE,changes <= tbins[(j-1)]),grep(TRUE,changes > tbins[j])))
        bin.crossers<-intersect(grep(TRUE,fos >= tbins[(j-1)]),grep(TRUE,los <= tbins[j])) # Taxa that pass straight through bin
        bin.bottoms<-intersect(intersect(grep(TRUE,fos >= tbins[(j-1)]),grep(TRUE,los > tbins[j])),grep(TRUE,los < tbins[(j-1)]))  # Taxa that span base of bin only
        bin.tops<-intersect(intersect(grep(TRUE,fos < tbins[(j-1)]),grep(TRUE,fos > tbins[j])),grep(TRUE,los <= tbins[j]))
        bin.within<-intersect(intersect(intersect(grep(TRUE,fos < tbins[(j-1)]),grep(TRUE,los > tbins[j])),grep(TRUE,fos > tbins[j])),grep(TRUE,los < tbins[(j-1)]))
        acctran.sumlineages[(j-1),i]<-(length(bin.crossers)*(tbins[(j-1)]-tbins[j]))+sum(tbins[(j-1)]-los[bin.bottoms])+sum(fos[bin.tops]-tbins[j])+sum(fos[bin.within]-los[bin.within])
    }
    cat(i," ")
}

###Stops working here - I think because too many NAs?

# Make character changes divided by sum of lineage durations table:
acctran.res<-acctranbins/acctran.sumlineages
deltran.res<-deltranbins/deltran.sumlineages
for(i in 1:ntrials) acctran.res[,i]<-as.numeric(gsub(NaN,0,acctran.res[,i]))
for(i in 1:ntrials) deltran.res[,i]<-as.numeric(gsub(NaN,0,deltran.res[,i]))

# Mean and CI vectors:
aclower95<-acctran.res[,1]
acmean95<-apply(acctran.res,1,mean)
acupper95<-acctran.res[,1]
delower95<-deltran.res[,1]
demean95<-apply(deltran.res,1,mean)
deupper95<-deltran.res[,1]

# Fill CIs:
for (i in 1:length(acctran.res[,1])) {
    aclower95[i]<-as.numeric(sort(acctran.res[i,])[max(floor(ntrials*0.025),1)])
    acupper95[i]<-as.numeric(sort(acctran.res[i,])[ceiling(ntrials*0.975)])
}
for (i in 1:length(deltran.res[,1])) {
    delower95[i]<-as.numeric(sort(deltran.res[i,])[max(floor(ntrials*0.025),1)])
    deupper95[i]<-as.numeric(sort(deltran.res[i,])[ceiling(ntrials*0.975)])
}

# Create cumulative character acquisition matrix:
acctranbinsper<-acctran.res
deltranbinsper<-deltran.res

# Fill it:
for (i in 1:ntrials) {
    for (j in 2:length(acctranbinsper[,1])) {
        acctranbinsper[j,i]<-acctranbinsper[j,i]+acctranbinsper[(j-1),i]
    }
}
for (i in 1:ntrials) {
    for (j in 2:length(deltranbinsper[,1])) {
        deltranbinsper[j,i]<-deltranbinsper[j,i]+deltranbinsper[(j-1),i]
    }
}

# Mean and CI vectors:
acclower95<-acctranbinsper[,1]
accmean95<-apply(acctranbinsper,1,mean)
accupper95<-acctranbinsper[,1]
declower95<-deltranbinsper[,1]
decmean95<-apply(deltranbinsper,1,mean)
decupper95<-deltranbinsper[,1]

# Fill CIs:
for (i in 1:length(acctranbins[,1])) {
    acclower95[i]<-as.numeric(sort(acctranbinsper[i,])[max(floor(ntrials*0.025),1)])
    accupper95[i]<-as.numeric(sort(acctranbinsper[i,])[ceiling(ntrials*0.975)])
}
for (i in 1:length(deltranbins[,1])) {
    declower95[i]<-as.numeric(sort(deltranbinsper[i,])[max(floor(ntrials*0.025),1)])
    decupper95[i]<-as.numeric(sort(deltranbinsper[i,])[ceiling(ntrials*0.975)])
}


# Combine on single plot:
pdf(file="charvstimeacctran.pdf",paper="a4r",width=30,height=20)
GeoPeriods=c(541,485.4,443.4, 419.2, 358.9,298.9,252.17)
par(mar=c(4,4,4,4)+0.1)
plot(x,accmean95,xlim=c(max(tbins),min(tbins)),type="l",lwd=1,axes=F,ylab="",ylim=c(0,max(accupper95)),xlab="", xaxs="i")
polygon(x=c(x,rev(x)),y=c(accupper95,rev(acclower95)),col="grey",border=NA, xaxs="i")
points(x,accmean95,type="l")
#points(x,acupper95,xlim=c(541,252.17),type="l",lwd=1,col="grey")
#points(x,aclower95,xlim=c(541,252.17),type="l",lwd=1,col="grey")
axis(4, ylab="Cumulative character acquisition")
mtext("Cumulative character acquisition",side=4, line=2.5)
par(new=TRUE)
plotCI(x,acmean95,uiw=(acupper95-acmean95), xaxt='n', xaxs="i", liw=(acmean95-aclower95),cex=0.5,xlim=c(max(GeoPeriods),min(GeoPeriods)),xlab="Time (Ma)",ylab="Per lineage duration character changes")
axis(1, at=c(500, 400, 450, 300, 350), lwd.ticks=2, xlim=c(max(GeoPeriods),min(GeoPeriods)))
axis(1, at=c(540,530,520,510,490,480,470,460,450,440,430,420,410,390,380,370,360,350,340,330,320,310,290,280,270,260), xlim=c(max(GeoPeriods),min(GeoPeriods)), labels=FALSE)
par(new=TRUE)
stratbar<-c(1,1,1,1,1, 1)
barwidths<-c(55.6, 42, 24.2, 60.3, 60, 46.73)
yvect=c(0.5)
barages<-c("Cambrian","Ordovician","Silurian","Devonian","Carboniferous","Permian")
bp<-barplot(stratbar, width=barwidths, space=0, col=FALSE, yaxt='n', ylim=c(0, 40), xaxs="i")
text(bp, yvect, barages, cex=0.6)
dev.off()

pdf(file="charvstimedeltran.pdf",paper="a4r",width=30,height=20)
GeoPeriods=c(541,485.4,443.4, 419.2, 358.9,298.9,252.17)
par(mar=c(4,4,4,4)+0.1)
plot(x,decmean95,xlim=c(max(tbins),min(tbins)),type="l",lwd=1,axes=F,ylab="",ylim=c(0,max(decupper95)),xlab="", xaxs="i")
polygon(x=c(x,rev(x)),y=c(decupper95,rev(declower95)),col="grey",border=NA, xaxs="i")
points(x,decmean95,type="l")
#points(x,deupper95,xlim=c(541,252.17),type="l",lwd=1,col="grey")
#points(x,delower95,xlim=c(541,252.17),type="l",lwd=1,col="grey")
axis(4, ylab="Cumulative character acquisition")
mtext("Cumulative character acquisition",side=4, line=2.5)
par(new=TRUE)
plotCI(x,demean95,uiw=(deupper95-demean95), xaxt='n', xaxs="i", liw=(demean95-delower95),cex=0.5,xlim=c(max(GeoPeriods),min(GeoPeriods)),xlab="Time (Ma)",ylab="Per lineage duration character changes")
axis(1, at=c(500, 400, 450, 300, 350), lwd.ticks=2, xlim=c(max(GeoPeriods),min(GeoPeriods)))
axis(1, at=c(540,530,520,510,490,480,470,460,450,440,430,420,410,390,380,370,360,350,340,330,320,310,290,280,270,260), xlim=c(max(GeoPeriods),min(GeoPeriods)), labels=FALSE)
par(new=TRUE)
stratbar<-c(1,1,1,1,1, 1)
barwidths<-c(55.6, 42, 24.2, 60.3, 60, 46.73)
yvect=c(0.5)
barages<-c("Cambrian","Ordovician","Silurian","Devonian","Carboniferous","Permian")
bp<-barplot(stratbar, width=barwidths, space=0, col=FALSE, yaxt='n', ylim=c(0, 40), xaxs="i")
text(bp, yvect, barages, cex=0.6)
dev.off()


# Write out data:
write.table(deltran.res,"deltranres.txt",sep=",",row.names=F,col.names=F)
write.table(acctran.res,"acctranres.txt",sep=",",row.names=F,col.names=F)


```

plot(tree)
edgelabels(tree$edge.lengths)




Work out if I can do this in Claddis
Also be cool to try and plot rate per character


Try:  ChronoPhyloMorphospacePlot
      ChangesinBins
      PlotCharacterChanges
